C/O and Snowline Locations in Protoplanetary Disks: The Effect of Radial Drift and Viscous Gas Accretion

### 2015-10-03 tex/draft/ms_final_revised.tex
update

\begin{abstract}
%The C/O ratio is an important signature of giant planet atmospheric composition and disk chemistry. 
The C/O ratio is a defining feature of both gas giant atmospheric and protoplanetary disk chemistry. In disks, the C/O ratio is regulated by the presence of snowlines of major volatiles at different distances from the central star.
We explore the effect of radial drift of solids and viscous gas accretion onto the central star on the snowline locations of the main C and O carriers in a protoplanetary disk, H$_2$O, CO$_2$ and CO, and their consequences for the C/O ratio in gas and dust throughout the disk. We determine the snowline locations for a range of fixed initial particle sizes, in both an active and a passive disk. We find that grains with sizes $\sim$$0.5$ cm $\lesssim s \lesssim$ 7 m for a passive disk and $\sim$$0.001$ cm $\lesssim s \lesssim$ 7 m for an active disk desorb %instantaneously and [I think instantaneously is too strong]
at a %fixed 
size-dependent location in the disk, which is independent %regardless 
of the particle's initial position. The snowline radius decreases for larger particles, up to sizes of $\sim$7~m. %location moves inwards as the particle size increases, and the innermost snowlines are set by grains with initial size $s\sim7$ m. 
Compared to a static disk, we find that radial drift and gas accretion move the H$_2$O and CO$_2$ snowlines inwards by up to 60 \%, and the CO snowline by up to 50 \%. We thus determine an inner limit on the snowline locations when radial drift and gas accretion are accounted for. %Our results have direct implications on the C/O ratio in giant planet atmospheres that have formed in situ. [I don't think we discuss this enough to warrant it as a conclusion
\end{abstract}

\section{Introduction}

@@

For an active disk, the radial drift velocity has an additional term due to the radial movement of the gas \citep{birnstiel12}, i.e.
\begin{equation}
\label{eq:rdotact}
\dot{r} \approx -2 \eta \Omega_{\rm k} r \Big(\frac{\tau_{\rm s}}{1+\tau_{\rm s}^2}\Big) + \frac{\dot{r}_{\rm gas}}{1+\tau_{\rm s}^2},
\end{equation}
where $\dot{r}_{\rm gas}$ is the radial gas accretion velocity and can be expressed as (e.g., \citealt{fkr02})
\begin{equation}
\label{eq:vgas}
\dot{r}_{\rm gas} = - \frac{3}{\Sigma \sqrt{r}} \frac{\partial}{\partial r}(\nu \Sigma \sqrt{r}) 
\end{equation}
with $\Sigma$ from Equation (\ref{eq:Sigmaact}). For the active disk steady-state solution (see Section \ref{sec:disk}), $\dot{r}_{\rm gas}$ can be expressed more simply using the definition of the mass flux, $\dot{M}=-2 \pi r \dot{r}_{\rm gas} \Sigma$, with $\dot{M}$ fixed and $\Sigma$ obtained from Equation (\ref{eq:Mdot}).

@@

For our choice of parameters, the minimum snowline radii are: $r_{\rm{H_2O}} \approx 0.3$ AU for the passive disk, $r_{\rm{H_2O}} \approx 0.26$ AU for the active disk and $r_{\rm{H_2O}} \approx 0.63$ AU for the active disk accretionally heated; $r_{\rm{CO_2}} \approx 3.7$ AU for the passive disk, $r_{\rm{CO_2}} \approx 3.4$ AU for both active disks; $r_{\rm{CO}} \approx 30$ AU for the passive and both active disks. For comparison, $r_{\rm{H_2O}} \approx 0.67$ AU, $r_{\rm{CO_2}} \approx 8.6$ AU and $r_{\rm{CO}} \approx 59$ AU for the static disk. Radial drift and gas accretion therefore push the snowline locations by up to $\sim$$60$ \% for H$_2$O and CO$_2$, and by up to $\sim$$50$ \% for CO.  We note that the H$_2$O snowline in all disks is significantly closer to the host star compared with Solar system models, which place the H$_2$O snowline between $\sim$$2.7$ to $\sim$$3.1$ AU (\citealt{hayashi81}, \citealt{podolak04}, \citealt{martin12}). This is partially because we choose a colder disk model, as well as the fact that gas accretion rates decrease over time, moving the snowline location outwards (see also \citealt{garaud07} and Section \ref{sec:neglected}). Observations of the CO snowline in TW Hya \citep{qi13} have found its location at a disk midplane temperature of 17 K (at 30 AU for the TW Hya specific temperature profile). The inferred  desorption temperature corresponds to the CO desorption temperature in a static disk, or to desorption from very small grains in an active disk, i.e. from grains that are too small to drift substantially. This suggests that the outer TW Hya disk is dominated by small grains, since larger particles would push the snowline location inwards, and therefore to higher desorption temperatures. This may seem contradictory to observations of grain growth in disks in general and in TW Hya in particular \citep{wilner00}. However, recent observations have revealed that grain growth is concentrated to the inner disk \citep{perez12} and outer disk snowlines may therefore be close to the ones expected in a static disk.

>>>>

\begin{abstract}
%The C/O ratio is an important signature of giant planet atmospheric composition and disk chemistry. 
The C/O ratio is a defining feature of both gas giant atmospheric and protoplanetary disk chemistry. In disks, the C/O ratio is regulated by the presence of snowlines of major volatiles at different distances from the central star.
We explore the effect of radial drift of solids and viscous gas accretion onto the central star on the snowline locations of the main C and O carriers in a protoplanetary disk, H$_2$O, CO$_2$ and CO, and their consequences for the C/O ratio in gas and dust throughout the disk. We determine the snowline locations for a range of fixed initial particle sizes, in both an active and a passive disk. We find that grains with sizes $\sim$$0.5$ cm $\lesssim s \lesssim$ 7 m for a passive disk and $\sim$$0.001$ cm $\lesssim s \lesssim$ 7 m for an active disk desorb %instantaneously and [I think instantaneously is too strong]
at a %fixed 
size-dependent location in the disk, which is independent %regardless 
of the particle's initial position. The snowline radius decreases for larger particles, up to sizes of $\sim$7~m. %location moves inwards as the particle size increases, and the innermost snowlines are set by grains with initial size $s\sim7$ m. 
Compared to a static disk, we find that radial drift and gas accretion \textbf{in a steady-state accretionally heated disk} move the H$_2$O snowline inwards by up to \textbf{40 \%}, the CO$_2$ snowlines by up to 60 \%, and the CO snowline by up to 50 \%. We thus determine an inner limit on the snowline locations when radial drift and gas accretion are accounted for. %Our results have direct implications on the C/O ratio in giant planet atmospheres that have formed in situ. [I don't think we discuss this enough to warrant it as a conclusion
\end{abstract}

\section{Introduction}
\label{sec:intro}

@@

For an active disk, the radial drift velocity has an additional term due to the radial movement of the gas \citep{birnstiel12}, i.e.
\begin{equation}
\label{eq:rdotact}
\dot{r} \approx -2 \eta \Omega_{\rm k} r \Big(\frac{\tau_{\rm s}}{1+\tau_{\rm s}^2}\Big) + \frac{\dot{r}_{\rm gas}}{1+\tau_{\rm s}^2},
\end{equation}
where $\dot{r}_{\rm gas}$ is the radial gas accretion velocity and can be expressed as (e.g., \citealt{fkr02})
\begin{equation}
\label{eq:vgas}
\dot{r}_{\rm gas} = - \frac{3}{\Sigma \sqrt{r}} \frac{\partial}{\partial r}(\nu \Sigma \sqrt{r}) 
\end{equation}
with $\Sigma$ from Equation (\ref{eq:Sigmaact}). For the active disk steady-state solution (see Section \ref{sec:disk}), $\dot{r}_{\rm gas}$ can be expressed more simply using the definition of the mass flux, $\dot{M}=-2 \pi r \dot{r}_{\rm gas} \Sigma$, with $\dot{M}$ fixed and $\Sigma$ obtained from Equation (\ref{eq:Mdot}). \textbf{For our choice of parameters for both active disks, we have found that the radial flow of gas is always directed inward, in contrast with the model of \citet{alidib14} which assumes that the gas drifts outwards (see Section \ref{sec:intro}). We note that variations in our fiducial disk model parameters (e.g., $T$, $\Sigma$, $\dot{M}$) may cause the gas to flow outwards in the outer parts of the disk, specifically at the CO$_2$ and CO snowline locations. Since drifting particles larger than a few cm are only modestly affected by gas accretion, an outward gas flow would move the CO$_2$ and CO locations further away from the star only for the smallest particles in our model, which are well-coupled to the gas.}

@@

For our choice of parameters, the minimum snowline radii are: $r_{\rm{H_2O}} \approx 0.3$ AU for the passive disk, $r_{\rm{H_2O}} \approx 0.26$ AU for the active disk and $r_{\rm{H_2O}} \approx 0.63$ AU for the active disk accretionally heated; $r_{\rm{CO_2}} \approx 3.7$ AU for the passive disk, $r_{\rm{CO_2}} \approx 3.4$ AU for both active disks; $r_{\rm{CO}} \approx 30$ AU for the passive and both active disks. For comparison, $r_{\rm{H_2O}} \approx 0.67$ AU\footnote{\textbf{For the viscous disk, we calculated the static snowline location using the same temperature profile as that of the viscous disk, for consistency purposes. Thus $r_{\rm{H_2O}} \approx 0.98$ AU for the static disk in this scenario.}}, $r_{\rm{CO_2}} \approx 8.6$ AU and $r_{\rm{CO}} \approx 59$ AU for the static disk. \textbf{For the viscous disk model, which is the most realistic,} radial drift and gas accretion push the snowline locations inwards by up to \textbf{$\sim$$40$ \%} for H$_2$O, by up to  $\sim$$60$ \% for CO$_2$, and by up to $\sim$$50$ \% for CO.  We note that the H$_2$O snowline in all disks is significantly closer to the host star compared with Solar system models, which place the H$_2$O snowline between $\sim$$2.7$ to $\sim$$3.1$ AU (\citealt{hayashi81}, \citealt{podolak04}, \citealt{martin12}). This is partially because we choose a colder disk model, as well as the fact that gas accretion rates decrease over time, moving the snowline location outwards (see also \citealt{garaud07} and Section \ref{sec:neglected}). Observations of the CO snowline in TW Hya \citep{qi13} have found its location at a disk midplane temperature of 17 K (at 30 AU for the TW Hya specific temperature profile). The inferred  desorption temperature corresponds to the CO desorption temperature in a static disk, or to desorption from very small grains in an active disk, i.e. from grains that are too small to drift substantially. This suggests that the outer TW Hya disk is dominated by small grains, since larger particles would push the snowline location inwards, and therefore to higher desorption temperatures. This may seem contradictory to observations of grain growth in disks in general and in TW Hya in particular \citep{wilner00}. However, recent observations have revealed that grain growth is concentrated to the inner disk \citep{perez12} and outer disk snowlines may therefore be close to the ones expected in a static disk.
